Finding shortest and closest vectors in a lattice of Voronoi's first kind

### 2014-04-15 talk/presentation.tex
added to presentation

\input{../../../bib/mathbf-abbrevs.sty}

@@

\title{Finding short vectors in a lattice of Voronoi's first kind}
\author{Robby McKilliam and Alex Grant \\ 
\vspace{0.5cm}
Institute for Telecommunications Research \\ University of South Australia
}

@@

\section{Lattices and short vectors}

\frame{
\frametitle{Lattices and short vectors}
An $n$-dimensional \term{lattice} $\Lambda$ is a discrete set of vectors from $\reals^m$, $m \geq n$, given by
\[
  \Lambda = \{ x = b_1 u_1 + b_2u_2 + \dots + b_n u_n \mid u_1, \dots u_n \in \ints \},
\] 
where $b_1, \dots, b_n \in \reals^m$ are \term{basis vectors} of $\Lambda$.
}

\frame{
\begin{figure}[tp]
	\centering 
		\includegraphics[width=\linewidth]{preslatticefigures-1.mps}
		\caption{A 2-dimensional lattice.}
		\label{fig:latfig1}
\end{figure}
}

\frame{
\frametitle{Lattices and short vectors}
Those points in the lattice with smallest non-zero length are called \term{short vectors}.  That is, the short vectors have squared length
\[
\min_{x\in \Lambda \backslash \{ 0 \} } \| x \|^2.
\]
}

\frame{
\begin{figure}[tp]
	\centering 
		\includegraphics[width=\linewidth]{preslatticefigures-1.mps}
		\caption{A 2-dimensional lattice.}
		\label{fig:latfig2}
\end{figure}
}

\frame{
\begin{figure}[tp]
	\centering 
		\includegraphics[width=\linewidth]{preslatticefigures-3.mps}
		\caption{A 2-dimensional lattice.  There are 4 short vectors.}
		\label{fig:latfigandpacking1}
\end{figure}
}

\frame{
\frametitle{Lattices and short vectors}
\begin{itemize}
\item Computing a short vector is called the \term{shortest vector problem}.
\item Applications in cryptography and number theory.
\item NP-hard for arbitrary lattices.
\item Easier for specific lattices.
\item For example, short vectors are easy to find in the \term{root lattices} $\ints^n$, $A_n$ and $D_n$.
\item We will show that the problem is relatively easy to solve for lattices of \term{Voronoi's first kind}.
\end{itemize}
}

@@

\frame{
\begin{figure}[tp]
	\centering 
        \includegraphics{graphs-3.mps}
		\caption{A graph with 4 vertices and 4 weighted edges.}
		\label{fig:graphexample}
\end{figure}
}

@@

\frame{
\begin{figure}[tp]
	\centering 
        \includegraphics{graphs-4.mps}
		\caption{A graph with 4 vertices and 4 weighted edges.}
		\label{fig:graphexample2}
\end{figure}
}

\frame{
\begin{figure}[tp]
	\centering 
        \includegraphics{graphs-5.mps} 
		\caption{The cut $C = \{v_2\}$ and $\bar{C} = \{v_1,v_3,v_4 \}$}
		\label{fig:graphcut1}
\end{figure}
}

\frame{
\begin{figure}[tp]
	\centering 
        \includegraphics{graphs-6.mps}
		\caption{The cut $C = \{v_2\}$ and $\bar{C} = \{v_1,v_3,v_4 \}$ has weight 5.}
		\label{fig:graphcut2}
\end{figure}
}

\frame{
\begin{figure}[tp]
	\centering 
        \includegraphics{graphs-7.mps}
		\caption{The minimum cut $C = \{v_3,v_4\}$ and $\bar{C} = \{v_1,v_2 \}$}
		\label{fig:graphcut3}
\end{figure}
}

\frame{
\begin{figure}[tp]
	\centering 
        \includegraphics{graphs-8.mps}
		\caption{The minimum cut $C = \{v_3,v_4\}$ and $\bar{C} = \{v_1,v_2 \}$ has weight 2.}
		\label{fig:graphcut4}
\end{figure}
}

@@

\frame{
\begin{figure}[tp]
	\centering 
        \includegraphics{graphs-4.mps}
		\caption{We have seen this graph before!}
		%\label{fig:graphcut4}
\end{figure}
}

\frame{
\begin{figure}[tp]
	\centering 
        \includegraphics{graphs-8.mps}
		\caption{The minimum cut $C = \{v_3,v_4\}$ and $\bar{C} = \{v_1,v_2 \}$ has weight 2.}
		%\label{fig:graphcut4}
\end{figure}
}

\frame{
\frametitle{Another example}
The minimum cut corresponds with the short vectors
\[
b_1 + b_2 = [1,1,0]
\]
and 
\[
b_3 + b_4 = -b_1 - b_2 = [-1,-1,0]
\]
of squared length 2.
}

\section{What's next}
\frame{
\frametitle{What's next}

\begin{itemize}
\item Is it possible to efficiently decide whether a lattice is of Voronoi's first kind?
\item Is it possible to efficiently find an obtuse superbase if it exists?
\item Can a similar approach be taken to solve the \term{nearest lattice point problem}?
\end{itemize}

@@

% \frame{
% \frametitle{The nearest point problem}
% \begin{itemize}
% \item Given some vector $y \in \reals^m$ we wish to find a lattice point $x \in \Lambda$ such that the squared Euclidean distance
% \[
% \| y - x \|^2
% \] 
% is minimised.  
% \item Equivalently we can minimise
% \[
% \| y - \sum_{i=1}^{n+1} b_i u_i \|^2
% \]
% over integers $u_1,\dots,u_{n+1}$.
% \end{itemize}
% }

% \frame{
% \begin{theorem}
% Let $z \in \reals^{n+1}$ be a column vector such that $y = \sum_{i=1}^{n+1}b_i z_i$.  A nearest point $x \in \Lambda$ to $y$ is of the form
% \[
% x = \sum_{i=1}^{n+1} b_i (\floor{z_i} + u_i)
% \]
% where the $u_i \in \{-1, 0, 1\}$ and $\floor{z_i}$ denotes the largest integer smaller that $z_i$.
% \end{theorem}

% %So for a lattice of Voronoi's first kind, the nearest point problem can be turned into a quadratic minimisation problem over vectors in $u_i \in \{-1,0,1\}$ rather that $u_i \in \ints$.

% }

>>>>

\input{../mathbf-abbrevs.sty}

@@

\title{Finding shortest and closest vectors in a lattice of Voronoi's first kind}
\author{Robby McKilliam, Alex Grant, and I. Vaughan L. Clarkson \\ 
\vspace{0.5cm}
{\small
Institute for Telecommunications Research \\ University of South Australia 
\\ \vspace{0.2cm}
School of Information Technology and Electrical Engineering \\ The University of Queensland
}
}

@@

\section{Lattices}

\frame{
\frametitle{Lattices}
An $n$-dimensional \term{lattice} $\Lambda$ is a discrete set of vectors from $\reals^m$, $m \geq n$, given by
\[
  \Lambda = \{ b_1 u_1 + b_2u_2 + \dots + b_n u_n \mid u_1, \dots u_n \in \ints \},
\] 
where $b_1, \dots, b_n \in \reals^m$ are \term{basis vectors} of $\Lambda$.
}

\frame{
\begin{figure}[tp]
	\centering 
		\includegraphics[width=\linewidth]{figs/preslatticefigures-1.mps}
		\caption{A 2-dimensional lattice.}
		\label{fig:latfig1}
\end{figure}
}

\frame{
\frametitle{Short vectors}
Those lattice points with smallest non-zero length are called \term{short vectors}.  That is, the short vectors have squared length
\[
\min_{x\in \Lambda \backslash \{ 0 \} } \| x \|^2.
\]
}

\frame{
\begin{figure}[tp]
	\centering 
		\includegraphics[width=\linewidth]{figs/preslatticefigures-1.mps}
		\caption{A 2-dimensional lattice.}
		\label{fig:latfig2}
\end{figure}
}

\frame{
\begin{figure}[tp]
	\centering 
		\includegraphics[width=\linewidth]{figs/preslatticefigures-3.mps}
		\caption{A 2-dimensional lattice.  There are 4 short vectors.}
		\label{fig:latfigandpacking1}
\end{figure}
}

\frame{
\frametitle{The Voronoi cell}
}

\frame{
\frametitle{The shortest vector problem}
\begin{itemize}
\item Computing a short vector is called the \term{shortest vector problem}.
\item Applications in cryptography and number theory.
\item NP-hard for arbitrary lattices.
\item Easier for specific lattices.
\item For example, short vectors are easy to find in the \term{root lattices} $\ints^n$, $A_n$ and $D_n$.
\item We will show that the problem is relatively easy to solve for lattices of \term{Voronoi's first kind}.
\end{itemize}
}

@@

\frame{
\begin{figure}[tp]
	\centering 
        \includegraphics{figs/graphs-3.mps}
		\caption{A graph with 4 vertices and 4 weighted edges.}
		\label{fig:graphexample}
\end{figure}
}

@@

\frame{
\begin{figure}[tp]
	\centering 
        \includegraphics{figs/graphs-4.mps}
		\caption{A graph with 4 vertices and 4 weighted edges.}
		\label{fig:graphexample2}
\end{figure}
}

\frame{
\begin{figure}[tp]
	\centering 
        \includegraphics{figs/graphs-5.mps} 
		\caption{The cut $C = \{v_2\}$ and $\bar{C} = \{v_1,v_3,v_4 \}$}
		\label{fig:graphcut1}
\end{figure}
}

\frame{
\begin{figure}[tp]
	\centering 
        \includegraphics{figs/graphs-6.mps}
		\caption{The cut $C = \{v_2\}$ and $\bar{C} = \{v_1,v_3,v_4 \}$ has weight 5.}
		\label{fig:graphcut2}
\end{figure}
}

\frame{
\begin{figure}[tp]
	\centering 
        \includegraphics{figs/graphs-7.mps}
		\caption{The minimum cut $C = \{v_3,v_4\}$ and $\bar{C} = \{v_1,v_2 \}$}
		\label{fig:graphcut3}
\end{figure}
}

\frame{
\begin{figure}[tp]
	\centering 
        \includegraphics{figs/graphs-8.mps}
		\caption{The minimum cut $C = \{v_3,v_4\}$ and $\bar{C} = \{v_1,v_2 \}$ has weight 2.}
		\label{fig:graphcut4}
\end{figure}
}

@@

\frame{
\begin{figure}[tp]
	\centering 
        \includegraphics{figs/graphs-4.mps}
		\caption{We have seen this graph before!}
		%\label{fig:graphcut4}
\end{figure}
}

\frame{
\begin{figure}[tp]
	\centering 
        \includegraphics{figs/graphs-8.mps}
		\caption{The minimum cut $C = \{v_3,v_4\}$ and $\bar{C} = \{v_1,v_2 \}$ has weight 2.}
		%\label{fig:graphcut4}
\end{figure}
}

\frame{
\frametitle{An example}
The minimum cut corresponds with the short vectors
\[
b_1 + b_2 = [1,1,0]
\]
and 
\[
b_3 + b_4 = -b_1 - b_2 = [-1,-1,0]
\]
of squared length 2.
}

\frame{
\frametitle{Some questions we asked in 2012}

\begin{itemize}
\item Can we efficiently decide whether a lattice is of Voronoi's first kind?
\item Can we efficiently find an obtuse superbase if it exists?
\item Can a similar approach be taken to solve the \term{nearest lattice point problem}?
\end{itemize}

@@

\frame{
\frametitle{Some questions we asked in 2012}

\begin{itemize}
\item Can we \sout{efficiently} decide whether a lattice is of Voronoi's first kind? {\color{red}\textbf{Yes}}
\item Can we \sout{efficiently} find an obtuse superbase if it exists? {\color{red}\textbf{Yes}}
\item Can a similar approach be taken to solve the \term{closest lattice point problem}? {\color{red}\textbf{$O(n^4)$}}
\end{itemize}

\frame{
START TALKING ABOUT CLOSEST POINT ALGORITHM
}

\section{What now?}
\frame{
\frametitle{What now?}

\begin{itemize}
\item Can good codes or quantisers be constructed from lattices of Voronoi's first kind?
\item Signal processing applications, phase unwrapping, circular statistics, etc.
\end{itemize}
